---
title: "IRIS Classification"
author: "Rahul Fernandes"
date: "2023-09-08"
output: github_document
---

## Description
The IRIS Classification Using R repository conducts exploratory data analysis on the IRIS dataset and a Polynomial Support Vector Machine classification model is build to classify the three species of the Iris flower.

## Setting Up the Environment
```{r loading packages, message=FALSE}
library(datasets)
library(dplyr)
library(ggplot2)
library(GGally)
library(caret)
```

## Exploratory Data Analysis
```{r include=FALSE}
iris <- datasets::iris
```

#### Descriptive Statistics
Summary of Statistics for the whole Dataset
```{r echo=FALSE}
summary(iris)
```

Descriptive statistics for Sepal Length by Species
```{r echo=FALSE}
iris %>% 
  group_by(Species) %>% summarize(
    mean(Sepal.Length), sd(Sepal.Length)
    )
```

Descriptive statistics for Sepal Width by Species
```{r echo=FALSE}
iris %>% 
  group_by(Species) %>% summarize(
    mean(Sepal.Width), sd(Sepal.Width)
    )
```

Descriptive statistics for Petal Length by Species
```{r echo=FALSE}
iris %>% 
  group_by(Species) %>% summarize(
    mean(Petal.Length), sd(Petal.Length)
  )
```

Descriptive statistics for Petal Width by Species
```{r echo=FALSE}
iris %>% 
  group_by(Species) %>% summarize(
    mean(Petal.Width), sd(Petal.Width)
  )
```

From the above, the dataset has three species of Iris flowers, Setosa, Versicolor and Virginica. There are 50 samples of each of these species, hence the dataset is balanced. In terms of Sepal Length, Virginica has the highest mean making it generally bigger on average with respect to Sepal Length. In terms of Sepal Width, Setosa has the highest mean making it generally bigger on average with respect to Sepal Width. In terms of Petal Width and Length, Virginica has the highest mean making it roughly the biggest species among the three.

#### Relationship Between Sepal and Petal Features
```{r echo=FALSE}
ggplot(iris) +
  geom_point(aes(x=Sepal.Width, y=Sepal.Length), color="red") +
    labs(title = "Sepal Width vs Sepal Length")
```

```{r echo=FALSE}
ggplot(iris) +
  geom_point(aes(x=Petal.Width, y=Petal.Length), color="blue") +
  labs(title = "Petal Width vs Petal Length")
```

#### Relationship Between Features and Species
```{r echo=FALSE}
# Relationship Sepal and Species
ggplot(iris) +
  geom_point(aes(x=Sepal.Width, y=Sepal.Length, color=Species)) +
  facet_wrap(~Species) +
  labs(title = "Relationship Between Sepal and Species")
```

```{r echo=FALSE}
# Relationship Petal and Species
ggplot(iris) +
  geom_point(aes(x=Petal.Width, y=Petal.Length, color=Species)) +
  facet_wrap(~Species) +
  labs(title = "Relationship Between Petal and Species")
```

There is an overlap in Sepal features among the three Species and there is no overlap in Petal features among the three species.

```{r echo=FALSE}
ggpairs(iris,
        columns = 1:4,
        aes(color = Species),
        upper = list(continuous = "points", combo = "box_no_facet"),
        legend = 2,
        title = "Relationship Between Features by Species"
        )
```

From the above pair plot, there is almost no overlap between features of the Setosa species, whereas there is overlap over features between Versicolor and Virginica species.

## Inference From Analysis
It can be inferred that Virginica is generally the biggest among the three species of the Iris flower and Setosa is the smallest. The relationship between the features and species shows that there is an overlap of features between Versicolor and Virginica. The mean and standard deviation of the Sepal Width feature are similar among the three species. This might indicate that this feature might not be as important while trying to differentiate between the three species. The graph of relationship between features by species shows that the classification is not linear.

## Classification
#### Splitting Dataset
The dataset is split into two parts, training and testing, 80% of the dataset is used for training and the rest 20% is used for testing.
```{r}
set.seed(100)

TrainingIndex <- createDataPartition(iris$Species, p=0.8, list = FALSE)
TrainingSet <- iris[TrainingIndex,]
TestingSet <- iris[-TrainingIndex,]
```
#### Building the model
```{r}
# Build Training Model
svmModel <- train(Species ~ ., data = TrainingSet,
               method = "svmPoly",
               na.action = na.omit,
               preProcess=c("scale","center"),
               trControl= trainControl(method="none"),
               tuneGrid = data.frame(degree=1,scale=1,C=1)
               )
print(svmModel)
```

#### Testing the model against the testing data
```{r}
# Model performance
svmModel.testing <- predict(svmModel, TestingSet)
svmModel.testing.confusion <- confusionMatrix(svmModel.testing,
                                              TestingSet$Species
                                              )
# Print Confusion Matrix
print(svmModel.testing.confusion)
```

#### Feature Importance
```{r echo=FALSE}
# Feature importance
Importance <- varImp(svmModel)
plot(Importance, col = "red")
```
From the above graph, the feature Sepal Width has the least importance in the model when classifying the three species of the Iris flower. Hence, proving the hypothesis made during analysis of the dataset.

## Conclusion
A Polynomial Support Vector Machine classification model is trained on 80% of the IRIS Dataset and tested on 20% of the IRIS Dataset. The model was compared with KNN and Random Forest models and the SVM model showed the highest metrics among the other models with an accuracy of 96.67% in classifying the three species of the Iris flower.