---
title: "IRIS Classification"
author: "Rahul Fernandes"
date: "2023-09-08"
output: html_document
---

## Description

The IRIS Classification Using R repository conducts exploratory data analysis on the IRIS dataset and a Polynomial Support Vector Machine classification model is build to classify the three species of Iris flowers.

## Setting up the environment

Loading Requirements

```{r loading packages, message=FALSE}
library(datasets)
library(dplyr)
library(ggplot2)
library(GGally)
library(caret)
```

## Exploratory Data Analysis
```{r include=FALSE}
iris <- datasets::iris
```
#### Descriptive Statistics
Summary of Statistics for the whole Dataset
```{r echo=FALSE}
summary(iris)
```
Descriptive statistics for Sepal Length by Species
```{r echo=FALSE}
iris %>% 
  group_by(Species) %>% summarize(
    mean(Sepal.Length), sd(Sepal.Length), min(Sepal.Length), max(Sepal.Length)
    )
```
Descriptive statistics for Sepal Width by Species
```{r echo=FALSE}
iris %>% 
  group_by(Species) %>% summarize(
    mean(Sepal.Width), sd(Sepal.Width), min(Sepal.Width), max(Sepal.Width)
    )
```
Descriptive statistics for Petal Length by Species
```{r echo=FALSE}
iris %>% 
  group_by(Species) %>% summarize(
    mean(Petal.Length), sd(Petal.Length), min(Petal.Length), max(Petal.Length)
  )
```
Descriptive statistics for Petal Width by Species
```{r echo=FALSE}
iris %>% 
  group_by(Species) %>% summarize(
    mean(Petal.Width), sd(Petal.Width), min(Petal.Width), max(Petal.Width)
  )
```

#### Relationship Between Sepal and Petal Features
```{r echo=FALSE}
ggplot(iris) +
  geom_point(aes(x=Sepal.Width, y=Sepal.Length), color="red") +
    labs(title = "Sepal Width vs Sepal Length")
```

```{r echo=FALSE}
ggplot(iris) +
  geom_point(aes(x=Petal.Width, y=Petal.Length), color="blue") +
  labs(title = "Petal Width vs Petal Length")
```

#### Relationship Between Features and Species
```{r echo=FALSE}
# Relationship Sepal and Species
ggplot(iris) +
  geom_point(aes(x=Sepal.Width, y=Sepal.Length, color=Species)) +
  facet_wrap(~Species) +
  labs(title = "Relationship Between Sepal and Species")
```

```{r echo=FALSE}
# Relationship Petal and Species
ggplot(iris) +
  geom_point(aes(x=Petal.Width, y=Petal.Length, color=Species)) +
  facet_wrap(~Species) +
  labs(title = "Relationship Between Petal and Species")
```

```{r echo=FALSE}
ggpairs(iris,
        columns = 1:4,
        aes(color = Species),
        upper = list(continuous = "points", combo = "box_no_facet"),
        legend = 2,
        title = "Relationship Between Features by Species"
        )
```

## Classification
#### Splitting Dataset
The Dataset is split into two parts, training and testing, 80% of the dataset is used for training and the rest 20% is used for testing.
```{r}
set.seed(100)

TrainingIndex <- createDataPartition(iris$Species, p=0.8, list = FALSE)
TrainingSet <- iris[TrainingIndex,]
TestingSet <- iris[-TrainingIndex,]
```
#### Building the model
```{r}
# Build Training Model
svmModel <- train(Species ~ ., data = TrainingSet,
               method = "svmPoly",
               na.action = na.omit,
               preProcess=c("scale","center"),
               trControl= trainControl(method="none"),
               tuneGrid = data.frame(degree=1,scale=1,C=1)
               )
print(svmModel)
```

#### Testing the model against the testing data
```{r}
# Model performance
svmModel.testing <- predict(svmModel, TestingSet)
svmModel.testing.confusion <- confusionMatrix(svmModel.testing,
                                              TestingSet$Species
                                              )
# Print Confusion Matrix
print(svmModel.testing.confusion)
```

#### Feature Importance
```{r echo=FALSE}
# Feature importance
Importance <- varImp(svmModel)
plot(Importance, col = "red")
```